Consistency and Fault Tolerance Considerations for the Next Iteration of the DOE Fast Forward Storage and IO Project

### 2014-04-28 paper.tex
IASDS submission version

\addtolength{\parskip}{-0.06in}

@@

@@

The rest of the paper is organized as follows. Section~\ref{sec:burst}
discusses some of the features of incorporating burst buffers as designed and
suggests some considerations and alternatives for the next generation of this
project. Section~\ref{sec:transactions} discusses the transactions approach
offered in the IOD layer and the corresponding epochs in the DAOS layer. It
also offers a comparison to the \DDT system given the very similar high-level
design and motivating use case.  Section~\ref{sec:summary} discusses the system
overall with recommendations on what design elements should be considered based
on broader issues with current HPC data centers. The paper is concluded in
Section~\ref{sec:conclusion} with a summary of the broad issues covered in the
paper.

\section{Burst Buffers}
\label{sec:burst}

@@

One of the bigger concerns is the observation that the original data staging
proposals all used compute nodes while the newer proposals seek not only to
make them a fixed portion of the IO stack, but also shared across all machine
users. The PreDatA~\cite{zheng:2010:predata} paper in particular examines the
potential costs and advantages of where to place operators similar to the IOD
proposed function shipping. There are two key takeaways from PreDatA. First,
placement matters.  Depending on the communication intensity vs. computation
intensity, where along the IO path to place the operation can matter
significantly. Second, and more importantly, the amount of time spent
processing for the operators was stretched to the point where it consumed
nearly all of the time between IO operations. The given ratios of compute
processes to staging process examined is representative for future extreme
scale platforms. If anything, the ratios offer more staging processes than IOD
processes would be available.

@@

By concentrating the Burst Buffers and function shipping into the storage
stack, three problems arise.  First, the amount of network bandwidth, IO
bandwidth, and compute power consumed for example operations from a single
application is likely to completely monopolize the IOD processes. Second, if
space and time partitioning is used instead, the functionality risks being too
small to be useful. Third, the long-term hardware performance advantage for
SSDs is questionable. Recent studies have shown that the erase-before-write
and interference between reading and writing with flash-based SSDs can cause
severe performance problems~\cite{skourtis:2013:ssd-performance}. The inclusion
of an optional use SSD layer in the new Trinity machine at Los Alamos will
offer a test bed to determine how likely these observed problems would affect a
production extreme scale platform.

%Current NAND-based flash devices top out at around 400 MB/sec.
%The key spec that is missing from this number is that 400 MB/sec is a measure
%of the fixed number of available IOPS multiplied by the block size. This
%represents the ideal streaming performance possible. The problem is that it
%costs an IOP to read 1 byte or 1 block (4KB or 8KB, depending on the device).
%It costs 1 IOP to write a full block--usually. In some cases, it will cost 2
%IOPs.  This accounts for the required pre-erase write prior to writing to a
%reused block. In the worst case, it can be 3 IOPS per write. This would be for
%a partial block write (read the old block, erase, write the modified block).
%One-third of 400 MB/sec, about 133 MB/sec, is well below the streaming
%performance of HDDs.  Granted, there is still rotational and seek latency to
%deal with for HDDs, but the advantage for SSDs has evaporated and potentially
%turned into a penalty at a considerable cost premium.  There are faster SSD
%solutions on the market that incorporate DRAM for caching and using the PCIe
%bus, for example, but their price precludes them from use in an extreme scale
%platform.

@@

\section{Transactions and Epochs}
\label{sec:transactions}

@@

\subsection{IOD Transactions}
To understand how transactions are used in the IOD layer, some terminology and
concepts must be explained first. At the coarsest grain level is a container.
Each container provides the single access context through which to access a
collection of objects. Transactions are the way that a series of modifications
to the objects within a container are treated atomically. Conceptually,
containers corresponds to a something akin to an HDF5 file in a traditional
file system. The objects in each container represent different data within a
file.  The three initially defined object types are key-value stores,
multi-dimensional arrays, and blobs.  The easiest way to understand these types
is to evaluate these from the perspective of an HDF-5 file, the initial user
interface layer. The key-value store represents a collection of attributes or
groups. The array represents a potentially multi-dimensional array.  The blob
represents a byte stream of arbitrary contents.  The fundamental difference
between an array and a blob is that the array has metadata specifying the
dimension(s). At the physical layer within the IO nodes, all of these objects
may be striped across multiple IO nodes.  Given this context, the transactions
come in two forms.

First is a single leader transaction where the IOD manages based on calls from
a single client. The underlying assumption is that the client side will manage
the transactional operations itself and the single client is capable of
reporting to the IOD how to evolve the transaction state.

The second form is called multi-leader and has the IOD layer manage the
transactions. In this case, when the transaction is created, a count of clients
is provided to the IOD layer. As clients commit their changes to the container,
the reference count is reduced. Once the count reaches 0, the transaction is
automatically committed.

@@

@@

The example metadata and data storage services created as part of the \DDT
project have no dependencies between transactions that prevent visibility
should an older version be incomplete. This additional, intentional requirement
by IOD offers different functionality than \DDTns's example services. In the
case of \DDTns, the functionality is more minimal, but also avoids some of the
concerns outlined below.

The second iteration of the protocol~\cite{lofstead:2013:pdsw-txn} fixed
scalability issues and demonstrated a scalable client-side coordination model
with excellent performance. The performance measured for a complex transaction
with \DDT is illustrated in Figure~\ref{fig:performance}. This performance is
explored in detail in a previous paper~\cite{lofstead:2013:pdsw-txn}.
%The
%breakdown of the number of participants in each role is shown in
%Table~\ref{tab:scaling}. For comparison, consider the Number of
%Sub-Coordinators equivalent to IOD processes. The Processes Per Sub-Coordinator
%represents the number of clients that use a particular IOD process. For these
%tests, we maintained a balanced distribution and always used at least two
%sub-coordinators to slow down the processing.

\begin{figure}[ht]
%\vspace{-0.15in}
\centering
\includegraphics[keepaspectratio=true, width=0.9\columnwidth]{images/performance}
\vspace{-0.15in}
\caption{Total Transaction Overhead}
\label{fig:performance}
%\vspace{-0.15in}
\end{figure}

%\begin{table}[ht]
%    \vspace{-0.15in}
%    \centering
%    \caption[Scaling Configuration]{Performance Tests Scaling Configuration}
%    \bigskip
%    \vspace{-0.15in}
%
%    \begin{tabular}{|r|r|r|}
%\hline
%Processes & \vtop{\hbox{\strut Number of}\hbox{\strut Sub-Coordinators}} & \vtop{\hbox{\strut Processes Per} \hbox{\strut Sub-Coordinator}}\\
%\hline
%%8 & 2 & 4 \\
%%16 & 2 & 8 \\
%%32 & 2 & 16 \\
%%64 & 2 & 32 \\
%%128 & 2 & 64 \\
%256 & 2 & 128 \\
%512 & 2 & 256 \\
%1024 & 4 & 256 \\
%2048 & 8 & 256 \\
%4096 & 16 & 256 \\
%8192 & 32 & 256 \\
%16384 & 64 & 256 \\
%32768 & 128 & 256 \\
%65536 & 256 & 256 \\
%\hline
%    \end{tabular}
%    \label{tab:scaling}
%\end{table}

@@

\begin{figure}[ht]
\centering
\includegraphics[keepaspectratio=true, width=0.5\columnwidth]{images/optimized-protocol}
\vspace{-0.15in}
\caption{Optimized Protocol}
\label{fig:optimized-protocol}
%\vspace{-0.25in}
\end{figure}

@@

%\subsection{Comparison to Other Protocols}
%Alternatives, such as Paxos~\cite{Lamport:1998:paxos} algorithms like
%ZooKeeper~\cite{Hunt:2010:zookeeper}, suffer from two limitations making them
%unsuitable for this environment. First, the distributed servers in Paxos
%systems are all distributed copies of each other that eventually become
%consistent. Given the scale we wish to address, a single node's memory is
%unlikely to be able to hold all of the data necessary for many operations at
%scale. They also do not have a guarantee for when consensus will be achieved
%without using slower synchronous calls. For the tight timing we wish to
%support, we need guarantees of when a consistent state has been achieved.
%Second, these systems also all assume that updates are initiated from a single
%client process rather than a parallel set of processes as is the standard in
%HPC environments.
%
%\DDT uses a second layer of coordination on the client side that greatly
%increases the scalability by consolidating messages from clients into unique
%sets prior to sending to the overall coordinator. A gossip
%protocol~\cite{ganesh:2003:gossip-protocols} may appear sufficient for this
%purpose, but the delay of eventual consistency is strictly avoided with this
%protocol to ensure guarantees at particular states in the code. For example, if
%a failure occurs, the global understanding of the role of all processes is
%required in order for effective communication to occur for operations like
%creating sub-transactions or voting. In this case, the protocol can offer
%stronger statements about consistency than these protocols offer.  These
%features offer a way to easily scale the transaction protocol given the
%guarantees we wish to offer. The IOD approach does nothing to address these
%concerns.
%
%Another effort to offer consistency and data integrity for the ZFS file
%system~\cite{zhang:2010:zfs} covers some of the same territory. Instead of a
%focus on the processes all having a notion of completion as a transaction, this
%work focuses on the integrity of the data movement operations. We view this
%work as something that should be considered hand-in-hand with a transaction
%approach to ensure the integrity of the movement of the data in addition to the
%agreement of processes about the successful completion of a parallel operation.

@@

>>>>

% change the inter-paragraph spacing
%\addtolength{\parskip}{-0.06in}

@@

The overall target machine architecture is illustrated in
Figure~\ref{fig:architecture}.

\begin{figure}[ht]
%\vspace{-0.15in}
\centering
\includegraphics[keepaspectratio=true, width=\columnwidth]{images/architecture}
%\vspace{-0.15in}
\caption{Target Machine Architecture}
\label{fig:architecture}
%\vspace{-0.15in}
\end{figure}

@@

The rest of the paper is organized as follows. First, an overview of the
target machine architecture and how it maps to the proposed components is
presented in Section~\ref{sec:overview}. Section~\ref{sec:burst} provides a
more detailed overview of the IOD layer and discusses some of the features of
incorporating burst buffers as designed and suggests some considerations and
alternatives for the next generation of this project.
Section~\ref{sec:transactions} discusses the transactions approach offered in
the IOD layer and the corresponding epochs in the DAOS layer. It also offers a
comparison to the \DDT system given the very similar high-level design and
motivating use case.  Section~\ref{sec:summary} discusses the system overall
with recommendations on what design elements should be considered based on
broader issues with current HPC data centers. The paper is concluded in
Section~\ref{sec:conclusion} with a summary of the broad issues covered in the
paper.

\section{Overview}
\label{sec:overview}

The demonstration system uses some particular existing technology to enable
testing the proposed concepts. The mapping is presented in
Figure~\ref{fig:arch-mapping}. We examine this mapping by evaluating the
descriptive blocks at the bottom of the figure.

\begin{figure}[ht]
%\vspace{-0.15in}
\centering
\includegraphics[keepaspectratio=true, width=\columnwidth]{images/arch-mapping}
%\vspace{-0.15in}
\caption{Mapping FFSIO Components to Machine Architecture}
\label{fig:arch-mapping}
%\vspace{-0.15in}
\end{figure}

The dark blue block that includes the HDF5 VOL block represents the User API
layer. It is worth noting that while this is the intended interaction
mechanism, it is still possible for other interfaces like MPI-IO and POSIX to
interact with the system. The IO Forwarding layer is represented by the two
black boxes. The represented split shows that the IO nodes can offer some
control over how data is moved from the compute area towards storage. The
IO dispatcher, the primary focus of the evaluation presented in this paper, is
shown in the light green box. The one extension is the dark pink block
representing technology like SSDs on the IO nodes. This is where a burst
buffer, if the NVRAM is available, would be hosted. The DAOS layer is the
yellow box. The VOSD is the purple box.

The specification of HDF5 and Lustre at the top and bottom of the stack is 
an expedience for demonstrating the functionality. In a production version of
the proposed system, these may be represented by other technologies.

Throughout the paper, multiple terms are used to define an operation applied to
data outside of the main application computation. These include both function
shipping and just the term operator. In all cases, the idea is a piece of code
is placed at some location along the IO path processing data as it passes
through that point. Note that this works in both directions as well as in
place. For example, when reading, an operator may be used to only retrieve the
pieces of a large data set from the DAOS and VOSD layers into the IOD caches
automatically.  It is also possible that an operator is deployed to either
rearrange data or generate a new copy with different ordering. For example,
changing the fast dimension for a multi-dimensional array can greatly improve
reading performance for particular reading patterns. In all cases, these
operators offer a chance to apply a relatively light piece of processing to
data rather than an intensive data analytics operation. These heavy-weight
operations should be left to the main computation resources leaving capacity
for other concurrent applications using the IOD layer.

\section{IOD Layer and Burst Buffers}
\label{sec:burst}

The IOD layer is intended to function as a higher performance, machine local
caching layer for the DAOS/VOSD layers. Even in a simplified system, by keeping
the IOD layer as part of the stack, making portable code is simpler since it
eliminates the need to use a different lower level APIs from the user layer
(e.g., HDF5) if the system lacks a feature, such as burst buffers. In an ideal
design, the IOD API could function solely an directly on the DAOS servers. The
reality is that performance demands and budget constraints will force something
like burst buffers to compensate for the performance afforded and to offer a
location to process data down to volumes that can fit through the limited
bandwidth. Given the criticality of burst buffers and the transaction mechanism
for end-to-end performance, these ideas are examined in detail below with
comparison to alternatives that may have been avoided due to perceived
performance limitations. In particular, the \DDT protocol was designed to
address these sorts of scenarios and has demonstrated excellent performance.

@@

One of the bigger concerns is that the original data staging proposals all used
compute nodes while the newer proposals seek not only to make them a fixed
portion of the IO stack, but also shared across all machine users. The
PreDatA~\cite{zheng:2010:predata} paper in particular examines the potential
costs and advantages of where to place operators similar to the IOD proposed
function shipping. There are two key takeaways from PreDatA. First, placement
matters.  Depending on the communication intensity vs. computation intensity,
where along the IO path to place the operation can matter significantly.
Second, and more importantly, the amount of time spent processing for the
operators was stretched to the point where it consumed nearly all of the time
between IO operations. The given ratios of compute processes to staging process
examined is representative for future extreme scale platforms. If anything, the
ratios offer more staging processes than IOD processes would be available.

@@

By concentrating the Burst Buffers and function shipping into the storage
stack, three problems arise.  First, the amount of network bandwidth, IO
bandwidth, and compute power consumed for example operations from a single
application is likely to completely monopolize the IOD processes. Second, if
space and time partitioning is used instead, the functionality risks being too
small to be useful. For example, if some of the IO nodes are dedicated either
throughout an application run (space partitioning) or for just phases of the
application runtime (time partitioning), the allocated capacity or ownership
duration may be insufficient to be useful. Third, the long-term hardware
performance advantage for SSDs is questionable. Recent studies have shown that
the erase-before-write and interference between reading and writing with
flash-based SSDs can cause severe performance
problems~\cite{skourtis:2013:ssd-performance}. The inclusion of an optional use
SSD layer in the new Trinity machine at Los Alamos will offer a test bed to
determine how likely these observed problems would affect a production extreme
scale platform.

Current NAND-based flash devices top out at around 400 MB/sec.  The key spec
that is missing from this number is that 400 MB/sec is a measure of the fixed
number of available IOPS multiplied by the block size. This represents the
ideal streaming performance possible. The problem is that it costs an IOP to
read 1 byte or 1 block (4KB or 8KB, depending on the device).  It costs 1 IOP
to write a full block--usually. In some cases, it will cost 2 IOPs.  This
accounts for the required pre-erase write prior to writing to a reused block.
In the worst case, it can be 3 or more IOPS per write. For example, if writing
forces both garbage collection and block compaction, many blocks may have to be
read, the data combined, the blocks erased, the combined data written, and then
the intended write operation can finally proceed on the newly freed blocks
(erase and then write).  One-third of 400 MB/sec, about 133 MB/sec, is well
below the streaming performance of HDDs.  Granted, there is still rotational
and seek latency to deal with for HDDs, but the advantage for SSDs has
evaporated and potentially turned into a considerable penalty at a cost
premium.  There are faster SSD solutions on the market that incorporate DRAM
for caching and using the PCIe bus, for example, but their price precludes them
from use in an extreme scale platform given budget constraints.

@@

\section{Transactions, Epochs, and Metadata}
\label{sec:transactions}

@@

\subsection{IOD Transactions}
To understand how transactions are used in the IOD layer, some terminology and
concepts must be explained first. At the coarsest grain level is a container.
Each container provides the single access context through which to access a
collection of objects. Transactions are used to treat a series of modifications
to the objects within a container atomically. Conceptually, containers
corresponds to a something akin to an HDF5 file in a traditional file system.
The objects in each container represent different data within a file.  The
three initially defined object types are key-value stores, multi-dimensional
arrays, and blobs.  The easiest way to understand these types is to evaluate
these from the perspective of an HDF-5 file, the initial user interface layer.
The key-value store represents a collection of attributes or groups. The array
represents a potentially multi-dimensional array.  The blob represents a byte
stream of arbitrary contents.  The fundamental difference between an array and
a blob is that the array has metadata specifying the dimension(s). At the
physical layer within the IO nodes, all of these objects may be striped across
multiple IO nodes.  Given this context, the transactions come in two forms.

First is a single leader transaction managed by IOD based on calls from a
single client. The underlying assumption is that the client side will manage
the transactional operations itself and the single client is capable of
reporting to the IOD how to evolve the transaction state. Passive failures are
detected based on the single client only. Any actively detected failure will
have to be managed by this single client.

The second form is called multi-leader and has the IOD layer manage the
transactions from a collection of clients. In this case, when the transaction
is created, a count of clients is provided to the IOD layer. As clients commit
their changes to the container, the reference count is reduced. Once the count
reaches 0, the transaction is automatically committed. Failure detection is
fully passive based on all participating clients.

@@

\subsection{Comparison to Other Protocols}
Alternatives, such as Paxos~\cite{Lamport:1998:paxos} algorithms like
ZooKeeper~\cite{Hunt:2010:zookeeper}, suffer from two limitations making them
unsuitable for this environment. First, the distributed servers in Paxos
systems are all distributed copies of each other that eventually become
consistent. Given the scale we wish to address, a single node's memory is
unlikely to be able to hold all of the data necessary for many operations at
scale. They also do not have a guarantee for when consensus will be achieved
without using slower synchronous calls. For the tight timing we wish to
support, we need guarantees of when a consistent state has been achieved.
Second, these systems also all assume that updates are initiated from a single
client process rather than a parallel set of processes as is the standard in
HPC environments.

Another effort to offer consistency and data integrity for the ZFS file
system~\cite{zhang:2010:zfs} covers some of the same territory. Instead of a
focus on the processes all having a notion of completion as a transaction, this
work focuses on the integrity of the data movement operations. We view this
work as something that should be considered hand-in-hand with a transaction
approach to ensure the integrity of the movement of the data in addition to the
agreement of processes about the successful completion of a parallel operation.

The transactions and epochs approach offered by FFSIO is most similar to
\DDT, but is a specialized implementation of a similar protocol. In this case,
\DDT is general and must use more general techniques. The FFSIO transactions
and epochs take advantage of simplifications based on the architecture and the
very specific use.

\DDT uses a second layer of coordination on the client side that greatly
increases the scalability by consolidating messages from clients into unique
sets prior to sending to the overall coordinator. A gossip
protocol~\cite{ganesh:2003:gossip-protocols} may appear sufficient for this
purpose, but the delay of eventual consistency is strictly avoided with this
protocol to ensure guarantees at particular states in the code. For example, if
a failure occurs, the global understanding of the role of all processes is
required in order for effective communication to occur for operations like
creating sub-transactions or voting. In this case, the protocol can offer
stronger statements about consistency than these protocols offer.  These
features offer a way to easily scale the transaction protocol given the
guarantees we wish to offer. The IOD approach does nothing to address these
concerns.

The details of \DDT is discussed next to offer a performance evaluation that
may be slower than FFSIO transactions and epochs for synchronous due to the
explicit messaging rather than relying on the hardware mechanisms. At scale,
the FFSIO performance should be similar to the \DDT approach.

@@

The example metadata and data storage services created as part of the \DDT
project have no dependencies between transactions that prevent visibility
should an older version be incomplete. This additional, intentional requirement
by IOD offers different functionality than \DDTns's example services. In the
case of \DDTns, the functionality is less, but also avoids some of the concerns
outlined below.

The second iteration of the protocol~\cite{lofstead:2013:pdsw-txn} fixed
scalability issues and demonstrated a scalable client-side coordination model
with excellent performance. The performance measured for a complex transaction
with \DDT is illustrated in Figure~\ref{fig:performance}. This performance is
explored in detail in a previous paper~\cite{lofstead:2013:pdsw-txn}.
The
breakdown of the number of participants in each role is shown in
Table~\ref{tab:scaling}. For comparison, consider the Number of
Sub-Coordinators equivalent to IOD processes. The Processes Per Sub-Coordinator
represents the number of clients that use a particular IOD process. For these
tests, we maintained a balanced distribution and always used at least two
sub-coordinators to slow down the processing.

\begin{figure}[ht]
%\vspace{-0.15in}
\centering
\includegraphics[keepaspectratio=true, width=\columnwidth]{images/performance}
%\vspace{-0.15in}
\caption{Total Transaction Overhead}
\label{fig:performance}
%\vspace{-0.15in}
\end{figure}

\begin{table}[ht]
    %\vspace{-0.15in}
    \centering
    \caption[Scaling Configuration]{Performance Tests Scaling Configuration}
    \bigskip
    %\vspace{-0.15in}

    \begin{tabular}{|r|r|r|}
\hline
Processes & \vtop{\hbox{\strut Number of}\hbox{\strut Sub-Coordinators}} & \vtop{\hbox{\strut Processes Per} \hbox{\strut Sub-Coordinator}}\\
\hline
%8 & 2 & 4 \\
%16 & 2 & 8 \\
%32 & 2 & 16 \\
%64 & 2 & 32 \\
%128 & 2 & 64 \\
256 & 2 & 128 \\
512 & 2 & 256 \\
1024 & 4 & 256 \\
2048 & 8 & 256 \\
4096 & 16 & 256 \\
8192 & 32 & 256 \\
16384 & 64 & 256 \\
32768 & 128 & 256 \\
65536 & 256 & 256 \\
\hline
    \end{tabular}
    \label{tab:scaling}
\end{table}

@@

\begin{figure}[ht]
\centering
\includegraphics[keepaspectratio=true, width=0.8\columnwidth]{images/optimized-protocol}
%\vspace{-0.15in}
\caption{Optimized Protocol}
\label{fig:optimized-protocol}
%\vspace{-0.25in}
\end{figure}

@@

@@

We wish to thank Eric Barton for the some of the figures used to illustrate
the system architecture.
